\LARGE \bf Stochastic Uncertainty Propagation in Power System Dynamics using Measure-valued Proximal Recursions

### 2021-07-04 Papers/Stochastic_Uncertainty_Propagation_TPS/JKO_Power_Systems_TPS.tex
updating Sec. V

\author{Abhishek~Halder,~\IEEEmembership{Member,~IEEE,} 
%\author{Abhishek~Halder,~\IEEEmembership{Member,~IEEE,}
Xinbo~Geng,~\IEEEmembership{Member,~IEEE,}
        Kenneth~F. Caluya,
        and~Pegah~Ojaghi% <-this % stops a space
\thanks{A. Halder and K.F. Caluya are with the Department of Applied Mathematics, University of California, Santa Cruz, CA 95064, USA, {\texttt{\{ahalder,kcaluya\}@ucsc.edu}}.}% <-this % stops a space
\thanks{X. Geng is with the School of Electrical and Computer Engineering, Cornell University, Ithaca, NY 14850, USA, {\texttt{xg72@cornell.edu}}.} %
\thanks{P. Ojaghi is with the Department of Computer Science and Engineering, University of California, Santa Cruz, CA 95064, USA, {\texttt{pojaghi@ucsc.edu}}.}%
\thanks{This research was partially supported by the NSF award 1923278.}}% <-this % stops a space
%\thanks{Manuscript received April 19, 2005; revised August 26, 2015.}}
% The paper headers
%\markboth{Journal of \LaTeX\ Class Files,~Vol.~14, No.~8, August~2015}%

@@

\subsubsection{Notation and organization}
We use boldfaced small letters for vectors, and boldfaced capital letters for matrices. The set of natural numbers, real numbers and complex numbers are denoted as $\mathbb{N}$, $\mathbb{R}$, and $\mathbb{C}$, respectively. The symbol $\nabla_{\bm{x}}$ denotes the Euclidean gradient operator with respect to (w.r.t.) the vector $\bm{x}$. Thus, $\nabla_{\bm{x}}\cdot$ stands for the divergence, and $\Delta_{\bm{x}}$ stands for the Laplacian w.r.t. vector $\bm{x}$. We use $\langle\cdot,\cdot\rangle$ to denote the standard Euclidean inner product. The real and imaginary parts of a complex number $z$ are denoted via $\Re(z)$ and $\Im(z)$, respectively. The uniform probability distribution over an interval $[a,b]$ is denoted as ${\rm{Unif}}\left([a,b]\right)$. Likewise, the $n$ dimensional uniform probability distribution over $[a,b]^{n}$ is denoted as ${\rm{Unif}}\left([a,b]^{n}\right)$. The symbol $\propto$ denotes proportionality, $|\cdot|$ denotes the absolute value, $\|\cdot\|_{2}$ denotes the standard Euclidean 2-norm, $\otimes$ denotes the Kronecker product, $\odot$ and $\oslash$ respectively denote the elementwise (Hadamard) product and division, $\det(\cdot)$ stands for the determinant, and the subscript $_{\sharp}$ denotes the pushforward of a PDF via a map. The $n\times n$ identity and zero matrices are denoted as $\bm{I}_{n}$ and $\bm{0}_{n\times n}$, respectively.

@@

\subsection{Sample Path Dynamics}\label{subsec:samplepathdynamics}
We consider the coupled stochastic differential equations (SDEs) associated with the networked-reduced power systems model \cite[Ch. 7]{sauerpai1998}. Specifically, for a power network with $n$ generators, the stochastic dynamics for the $i$-th generator is given by the It\^{o} SDEs
\begin{subequations}
\begin{align}
&{\mathrm{d}}\theta_{i} = \omega_{i}\:{\mathrm{d}}t,\label{RotAngleGen}\\
&m_{i}\:{\mathrm{d}}\omega_{i} =	\left(\!P_{i} - \gamma_{i}\omega_{i} - \displaystyle\sum_{j=1}^{n} k_{ij}\sin\left(\theta_{i}-\theta_{j}-\varphi_{ij}\right)\!\right){\mathrm{d}}t \nonumber\\
&\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad+\sigma_{i}\:{\mathrm{d}}w_{i},
\end{align}
\label{ItoSDEcomponentlevel}
\end{subequations}
where the state variables are the rotor angles $\theta_{i}\in[0,2\pi)$ and the rotor angular velocities $\omega_{i}\in\mathbb{R}$, for $i\in\{1,\hdots,n\}$. The stochastic forcing is modeled through the standard Wiener process $w_{i}(t)$, and the diffusion coefficient $\sigma_{i}>0$ denotes the intensity of stochastic forcing at the $i$-th generator.

With the $i$-th generator, we associate its inertia $m_{i}>0$ and damping coefficient $\gamma_{i}>0$. The other parameters: the \emph{effective power input} $P_{i}$, the \emph{phase shift} $\varphi_{ij}\in[0,\frac{\pi}{2})$, and the \emph{coupling coefficients} $k_{ij}\geq 0$, depend on the network reduced admittance matrix $\bm{Y}\equiv[Y_{ij}]_{i,j=1}^{n}\in\mathbb{C}^{n\times n}$. Specifically,
\begin{subequations}
\begin{align}
 P_{i} &= P_{i}^{\text{mech}} - E_{i}^{2} \Re\left(Y_{ii}\right), \\
 \varphi_{ij} &= \begin{cases} -\arctan\left(\dfrac{\Re\left(Y_{ij}\right)}{\Im\left(Y_{ij}\right)}\right), & \text{if}\;i\neq j,\\
 0, & \text{otherwise},
 \end{cases}\\
 k_{ij} &= \begin{cases} E_{i}E_{j} \vert Y_{ij}\vert, & \text{if}\;i\neq j,\\
 0, & \text{otherwise},
 \end{cases}
\end{align}
\end{subequations}
where $P_{i}^{\text{mech}}$ is the mechanical power input, and $E_{i}$ is the internal voltage magnitude for generator $i$.

@@

\begin{remark}
We clarify here that the network-reduced model \eqref{ItoSDEcomponentlevel} is obtained from the so-called structure preserving power network model \cite{odun2012structure} with $n$ generators, after applying the Kron reduction, and therefore, has all-to-all connection topology. If the unreduced network has $n$ generators and $m$ buses, then the unreduced admittance matrix $\bm{Y}_{\rm{unreduced}} \in \mathbb{C}^{(n+m)\times(n+m)}$ satisfies
\begin{align}
\bm{Y}_{\rm{unreduced}} = .
\label{unreducedY}	
\end{align}
The network reduced admittance matrix $\bm{Y}\in\mathbb{C}^{n\times n}$ is the Schur complement
\[\bm{Y} = \bm{Y}_{\rm{unreduced}}.\]	
\end{remark}

@@

The corresponding macroscopic dynamics governing the flow of the joint state PDF $\rho(t,\bm{\theta},\bm{\omega})$ is given by a kinetic Fokker-Planck partial differential equation (PDE)
\begin{align}
\frac{\partial \rho}{\partial t} &= - \langle \bm{\omega},\nabla_{\bm{\theta}}\rho \rangle  + \nabla_{\bm{\omega}} \cdot \left( \rho \left( \bm{M}^{-1}\bm{\Gamma}\bm{\omega} + \bm{M}^{-1}\nabla_{\bm{\theta}}V(\bm{\theta}) \right.\right.\nonumber\\
&\qquad\qquad\qquad \left.\left. +\frac{1}{2} \bm{M}^{-1}\bm{\Sigma}\bm{\Sigma}^{\top}\bm{M}^{-1} \nabla_{\bm{\omega}} \log \rho \right) \right),
\label{KineticFPK}	
\end{align}
subject to the initial condition \eqref{FPKInitCond} and the joint parameter PDF $\rho_{\text{param}}$. A direct numerical solution of this PDE initial value problem using conventional discretization (e.g., finite difference) or function approximation techniques will not be scalable in general, as explained in Sec. \ref{subsubsec:technicalchallenges}. In the next Section, we discuss how a measure-valued variational recursion proposed in our recent works \cite{caluya2019ACC,caluya2019TAC,halder2020hopfield,caluya2021TAC,caluya2020reflected} can be employed to address this challenge.

@@

@@

\begin{IEEEbiography}[{\includegraphics[height=1.25in]{Abhishek-Halder.jpg}}]{Abhishek
Halder}
(S'10--M'14) is an Assistant Professor in the Department of Applied Mathematics,
and is an affiliated faculty in the Department of Electrical and Computer Engineering at
University of California, Santa Cruz. Before that he held postdoctoral
positions in the Department of Mechanical and Aerospace Engineering at
University of California, Irvine, and in the Department of Electrical
and Computer Engineering at Texas A\&M University. He obtained his
Bachelors and Masters from Indian Institute of Technology Kharagpur in
2008, and Ph.D. from Texas A\&M University in 2014, all in Aerospace
Engineering. His research interests are in stochastic systems, control, learning and optimization with application focus on large scale cyber-physical
systems.
\end{IEEEbiography}

@@

\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{XinboGeng.jpg}}]{Xinbo Geng} (S'13--M'20)
received his B.E. degree in Electrical Engineering from Tsinghua University, Beijing, China, in 2013. He received the M.S. and Ph.D. degrees from Department of Electrical and Computer Engineering, Texas A\&M University, College Station, TX, in 2015 and 2019, respectively.
He is a postdoctoral researcher jointly affiliated with Cornell University and Texas A\&M University since 2020.
He was a power system design and studies intern at National Renewable Energy Lab in 2016, a data analyst intern at ISO New England in 2017, a visiting student at UC Berkeley and MIT in 2018, and a visiting student at University of Wisconsin Madison in 2019.
His main research interest is decision making under uncertainties.
\end{IEEEbiography}

>>>>

\author{Abhishek~Halder,~\IEEEmembership{Senior Member,~IEEE,} 
%\author{Abhishek~Halder,~\IEEEmembership{Member,~IEEE,}
Xinbo~Geng,~\IEEEmembership{Member,~IEEE,}
        Kenneth~F. Caluya,
        and~Pegah~Ojaghi% <-this % stops a space
\thanks{A. Halder and K.F. Caluya are with the Department of Applied Mathematics, University of California, Santa Cruz, CA 95064, USA, {\texttt{\{ahalder,kcaluya\}@ucsc.edu}}.}% <-this % stops a space
\thanks{X. Geng is with the School of Electrical and Computer Engineering, Cornell University, Ithaca, NY 14850, USA, {\texttt{xg72@cornell.edu}}.} %
\thanks{P. Ojaghi is with the Department of Computer Science and Engineering, University of California, Santa Cruz, CA 95064, USA, {\texttt{pojaghi@ucsc.edu}}.}%
\thanks{This research was partially supported by the NSF award 1923278.}}% <-this % stops a space
%\thanks{Manuscript received April 19, 2005; revised August 26, 2015.}}
% The paper headers
%\markboth{Journal of \LaTeX\ Class Files,~Vol.~14, No.~8, August~2015}%

@@

\subsubsection{Notation and organization}
We use boldfaced small letters for vectors, and boldfaced capital letters for matrices. The set of natural numbers, real numbers and complex numbers are denoted as $\mathbb{N}$, $\mathbb{R}$, and $\mathbb{C}$, respectively. The symbol $\nabla_{\bm{x}}$ denotes the Euclidean gradient operator with respect to (w.r.t.) the vector $\bm{x}$. Thus, $\nabla_{\bm{x}}\cdot$ stands for the divergence, and $\Delta_{\bm{x}}$ stands for the Laplacian w.r.t. vector $\bm{x}$. We use $\langle\cdot,\cdot\rangle$ to denote the standard Euclidean inner product. The real and imaginary parts of a complex number $z$ are denoted via $\Re(z)$ and $\Im(z)$, respectively. We use the superscript $*$ to denote the complex conjugate. The uniform probability distribution over an interval $[a,b]$ is denoted as ${\rm{Unif}}\left([a,b]\right)$. Likewise, the $n$ dimensional uniform probability distribution over $[a,b]^{n}$ is denoted as ${\rm{Unif}}\left([a,b]^{n}\right)$. The symbol $\propto$ denotes proportionality, $|\cdot|$ denotes the absolute magnitude, $\|\cdot\|_{2}$ denotes the standard Euclidean 2-norm, $\otimes$ denotes the Kronecker product, $\odot$ and $\oslash$ respectively denote the elementwise (Hadamard) product and division, $\det(\cdot)$ stands for the determinant, and the subscript $_{\sharp}$ denotes the pushforward of a PDF via a map. The $n\times n$ identity and zero matrices are denoted as $\bm{I}_{n}$ and $\bm{0}_{n\times n}$, respectively.

@@

\subsection{Sample Path Dynamics}\label{subsec:samplepathdynamics}

%\subsubsection{Deterministic dynamics and network reduced model}
%Before considering stochastic uncertainties in the dynamics, let us briefly recall the network reduced power systems model \cite[Ch. 7]{sauerpai1998} in the deterministic case. Suppose we denote the collection of generator buses in a networked power system as $\mathcal{G}$, and the collection of non-generator buses as $\mathcal{N}$. Assuming that the non-generator buses are modeled as constant current loads, with currents $I_i$, $i \in \mathcal{N}$, the power flow equations become
%\begin{subequations}
%\begin{eqnarray}
%m_i \ddot{\theta} + \gamma_i \dot{\theta_i} = P_{mech,i} - P_{load,i} - \sum_{j} |V_i| |V_j| |Y_{ij}| \sin(\theta_i - \theta_j + \phi_{ij}) &~ i \in \mathcal{G}~\text{buses with generators} \\
%0 = - \text{Real}(V_i I_i^\dagger) - \sum_{j} |V_i| |V_j| |Y_{ij}| \sin(\theta_i - \theta_j + \phi_{ij}) &~ i \in \mathcal{L}~\text{buses without generators} \\
%\end{eqnarray}
%\end{subequations}
%Note: $P_{mech,i} \in \mathbb{R}$, $P_{load,i} \in \mathbb{R}$ and $I_i \in \mathbb{C}$ are all give parameters; load current is a complex number. 
%
%
%\subsubsection{Stochastic dynamics}
In this work, we consider the coupled stochastic differential equations (SDEs) associated with the networked-reduced power systems model \cite[Ch. 7]{sauerpai1998}. Specifically, for a power network with $n$ generators, the stochastic dynamics for the $i$-th generator is given by the It\^{o} SDEs
\begin{subequations}
\begin{align}
&{\mathrm{d}}\theta_{i} = \omega_{i}\:{\mathrm{d}}t,\label{RotAngleGen}\\
&m_{i}\:{\mathrm{d}}\omega_{i} =	\left(\!P_{i} - \gamma_{i}\omega_{i} - \displaystyle\sum_{j=1}^{n} k_{ij}\sin\left(\theta_{i}-\theta_{j}-\varphi_{ij}\right)\!\right){\mathrm{d}}t \nonumber\\
&\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad+\sigma_{i}\:{\mathrm{d}}w_{i},
\end{align}
\label{ItoSDEcomponentlevel}
\end{subequations}
where the state variables are the rotor angles $\theta_{i}\in[0,2\pi)$ and the rotor angular velocities $\omega_{i}\in\mathbb{R}$, for $i\in\{1,\hdots,n\}$. The stochastic forcing is modeled through the standard Wiener process $w_{i}(t)$, and the diffusion coefficient $\sigma_{i}>0$ denotes the intensity of stochastic forcing at the $i$-th generator.

\begin{remark}
We emphasize here that the network reduced model \eqref{ItoSDEcomponentlevel} is obtained from the so-called structure preserving power network model \cite{odun2012structure} after applying the Kron reduction \cite{dorfler2012kron}, and therefore, has all-to-all connection topology.	
\end{remark}

With the $i$-th generator, we associate its inertia $m_{i}>0$ and damping coefficient $\gamma_{i}>0$. The other parameters: the \emph{effective power input} $P_{i}$, the \emph{phase shift} $\varphi_{ij}\in[0,\frac{\pi}{2})$, and the \emph{coupling coefficients} $k_{ij}\geq 0$, depend on the network reduced admittance matrix $\bm{Y}\equiv[Y_{ij}]_{i,j=1}^{n}\in\mathbb{C}^{n\times n}$. Specifically,
\begin{subequations}
\begin{align}
 P_{i} &= P_{i}^{\text{mech}} - P_{i}^{\text{load}} - |V_{i}|^{2} \Re\left(Y_{ii}\right) + \Re\left(V_i \cdot I_{i}^{*}\right),\\
 \varphi_{ij} &= \begin{cases} -\arctan\left(\dfrac{\Re\left(Y_{ij}\right)}{\Im\left(Y_{ij}\right)}\right), & \text{if}\;i\neq j,\\
 0, & \text{otherwise},
 \end{cases}\\
 k_{ij} &= \begin{cases} |V_{i}| |V_{j}| \vert Y_{ij}\vert, & \text{if}\;i\neq j,\\
 0, & \text{otherwise},
 \end{cases}
\end{align}
\end{subequations}
where $P_{i}^{\text{mech}}$ is the mechanical power input, and $|V_{i}|$ is the voltage magnitude for generator $i$. {\red{{\textbf{Xinbo:}} One sentence saying any approximation made in (2) by using $V_{i}$ instead of the internal voltage $E_{i}$}}

@@

\begin{remark}
If the unreduced network has $n$ generators and $m$ buses, then the unreduced admittance matrix $\bm{Y}_{\rm{unreduced}} \in \mathbb{C}^{(n+m)\times(n+m)}$ satisfies
\begin{align}
\bm{Y}_{\rm{unreduced}} = .
\label{unreducedY}	
\end{align}
The network reduced admittance matrix $\bm{Y}\in\mathbb{C}^{n\times n}$ is the Schur complement
\[\bm{Y} = \bm{Y}_{\rm{unreduced}}.\]	
\end{remark}

@@

The corresponding macroscopic dynamics governing the flow of the joint state PDF $\rho(t,\bm{\theta},\bm{\omega})$ is given by a kinetic Fokker-Planck partial differential equation (PDE)
\begin{align}
\frac{\partial \rho}{\partial t} &= - \langle \bm{\omega},\nabla_{\bm{\theta}}\rho \rangle  + \nabla_{\bm{\omega}} \cdot \left( \rho \left( \bm{M}^{-1}\bm{\Gamma}\bm{\omega} + \bm{M}^{-1}\nabla_{\bm{\theta}}V(\bm{\theta}) \right.\right.\nonumber\\
&\qquad\qquad\qquad \left.\left. +\frac{1}{2} \bm{M}^{-1}\bm{\Sigma}\bm{\Sigma}^{\top}\bm{M}^{-1} \nabla_{\bm{\omega}} \log \rho \right) \right),
\label{KineticFPK}	
\end{align}
subject to the initial condition \eqref{FPKInitCond} and the joint parameter PDF $\rho_{\text{param}}$. Of course, this subsumes special cases when either the initial condition or the parameter vector is deterministic. A direct numerical solution of this PDE initial value problem using conventional discretization (e.g., finite difference) or function approximation techniques will not be scalable in general, as explained in Sec. \ref{subsubsec:technicalchallenges}. In the next Section, we discuss how a measure-valued variational recursion proposed in our recent works \cite{caluya2019ACC,caluya2019TAC,halder2020hopfield,caluya2021TAC,caluya2020reflected} can be employed to address this challenge.

@@

% === Kron reduction

\bibitem{dorfler2012kron}
F.~Dorfler, and F.~Bullo, ``Kron reduction of graphs with applications to electrical networks", \emph{IEEE Transactions on Circuits and Systems I: Regular Papers}, Vol. 60, No. 1, pp. 150--163, 2012.

@@

\begin{IEEEbiography}[{\includegraphics[height=1.25in]{Abhishek-Halder.jpg}}]{Abhishek
Halder}
(S'10-M'14-SM'21) is an Assistant Professor in the Department of Applied Mathematics,
and is an affiliated faculty in the Department of Electrical and Computer Engineering at
University of California, Santa Cruz. Before that he held postdoctoral
positions in the Department of Mechanical and Aerospace Engineering at
University of California, Irvine, and in the Department of Electrical
and Computer Engineering at Texas A\&M University. He obtained his
Bachelors and Masters from Indian Institute of Technology Kharagpur in
2008, and Ph.D. from Texas A\&M University in 2014, all in Aerospace
Engineering. His research interests are in stochastic systems, control, learning and optimization with application focus on large scale cyber-physical
systems.
\end{IEEEbiography}

@@

\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{XinboGeng.jpg}}]{Xinbo Geng} (S'13-M'20)
received his B.E. degree in Electrical Engineering from Tsinghua University, Beijing, China, in 2013. He received the M.S. and Ph.D. degrees from Department of Electrical and Computer Engineering, Texas A\&M University, College Station, TX, in 2015 and 2019, respectively.
He is a postdoctoral researcher jointly affiliated with Cornell University and Texas A\&M University since 2020.
He was a power system design and studies intern at National Renewable Energy Lab in 2016, a data analyst intern at ISO New England in 2017, a visiting student at UC Berkeley and MIT in 2018, and a visiting student at University of Wisconsin Madison in 2019.
His main research interest is decision making under uncertainties.
\end{IEEEbiography}
